---
title: 'ACMT Example: Calculating a Neighborhood Deprivation Index'
author: "Amy Youngbloom"
date: "1/13/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# knit this file with the working directory of "workspace/" instead of "workspace/examples"
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '../')
```

## Introduction
The ACMT can also be used to generate interpolated measures based on standard neighborhood indices that use American Community Survey data. These indices provide useful and consistent ways for researchers to operationalize measures such as deprivation, social vulnerability, or social fragmentation. One example of an index measure such as this is the Neighborhood Deprivation Index (NDI) developed by Messer et al. (2006). The NDI is a composite measure using socioeconomic variables to describe the level of deprivation of a given area. The NDI relies on American Community Survey variables, all of which are built into the default list of variables in the ACMT. As such the NDI may be a useful tool in comparing interpolated neighborhood deprivation levels around given points of interest.

In this example, we will utilize a list of establishments in Seattle either with an active marijuana retailer license or with an active alcohol license and use the NDI to compare levels of neighborhood depravity around each type of establishment.

We take the following steps:

  1. Download dispensary and alcohol retailer data and reformat into lists from WA Liquor and Cannabis Public Record website: https://lcb.wa.gov/records/frequently-requested-lists

  2. Geocode dispensary and alcohol retailers
  3. Designate the ACS measures relevant to the NDI measure
  4. Create a function and loop to pull ACMT measures
  5. Run a principal component analysis to find weights for each variable measure
  6. Calculate the NDI measure
  7. Create a boxplot to compare the NDI measures for dispensaries versus alcohol retailers
  8. Calculate a t-test to statistically compare NDI measures

```{r, results='hide'}
source("setup-acmt.R")
library(readxl)
library(tidyverse)
install.packages('janitor')
library(janitor)
```

## 1. Download data

Verify the urls below for the Dispensaries and Alcohol retailers below. The lists are updated regularly and can be found here: https://lcb.wa.gov/records/frequently-requested-lists. See urls for these lists: Cannabis License Applicants and Off Premises, Licensees

```{r, results='hide'}

#Find URL For most up-tp-date lists:
download.file(url = "https://lcb.wa.gov/sites/default/files/publications/Public_Records/2019/CannabisApplicants03222022.xls", 
              destfile = "external_data/downloaded_cannabis.xls") ##check records website if url is outdated


download.file(url = "https://lcb.wa.gov/sites/default/files/publications/Public_Records/2019/Off%20Premise03222022.xlsx", 
              destfile = "external_data/downloaded_alcohol.xlsx")

cannabis_dataframe <- read_excel("external_data/downloaded_cannabis.xls")
alcohol_dataframe<-read_excel("external_data/downloaded_alcohol.xlsx")

#limit to Seattle only dispensaries and create address field:
seattle_dispensary<-cannabis_dataframe %>%
  janitor::clean_names() %>%
  filter(city == "SEATTLE", priv_desc == "MARIJUANA RETAILER", (privilege_status=="ACTIVE (ISSUED)" | privilege_status == 'PENDING (ISSUED)')) %>%	
  mutate(full_address = paste(street_address, city, state, zip_code, sep = ", "), ID=paste(as.character(row_number()), 'disp', sep='.'), status=privilege_status, license_type='dispensary') %>%
  subset(select=c(ID, tradename, status, full_address, license_type)) %>%
  mutate(privilege='marijuana retailer')

seattle_alcohol<-alcohol_dataframe %>%
  janitor::clean_names() %>%
  filter(loc_city == "SEATTLE", status == "ACTIVE (ISSUED)" | status == "PENDING (ISSUED)", privilege != "BEER/WINE GIFT DELIVERY") %>%
  mutate(full_address = paste(loc_address, loc_city, loc_st, loc_zip, sep = ", "), 
        ID=paste(as.character(row_number()), 'alcohol', sep='.'), 
         license_type='alcohol_offpremise') %>%
        subset(select=c(ID, tradename, status, full_address, license_type, privilege))
## Limiting list for speed of processing
seattle_alcohol_dispensary_list<-rbind(seattle_dispensary[1:100,], seattle_alcohol[1:200,])
```

```{r}
head(seattle_alcohol_dispensary_list)
```

## 2. Write a geocoding loop to geocode all addresses

```{r, results='hide'}
#add lat and long variables
seattle_alcohol_dispensary_list<-seattle_alcohol_dispensary_list %>%
  filter(!is.na(full_address))%>%
  sample_n(100)%>% #subset to speed up the processing
mutate(lat=NA, 
       long=NA)
  

#Geocoding loop
for(i in 1:nrow(seattle_alcohol_dispensary_list)) {
  if(!is.na(seattle_alcohol_dispensary_list$lat[i])) next #skip already geocoded)
  if(is.na(seattle_alcohol_dispensary_list$full_address[i])) next #skip NA address values
  print(i) #print # so you know the progress of the geocoder
  address<-seattle_alcohol_dispensary_list$full_address
  lat_long<-geocode(address[i])
  seattle_alcohol_dispensary_list$lat[i]<-lat_long$latitude #add latitude to dataset
  seattle_alcohol_dispensary_list$long[i]<-lat_long$longitude #add long to dataset
}
```

## 3. Designate the ACS measures relevant to the NDI measure
Once the addresses are all geocoded, we need to designate which variables will be pulled and create ’_count’ and ’_proportion’ versions of each relevant variable. Note that ACS variable names change from year to year. These variable names set below are for 2019. For other years, refer to the ACS Variables by Year spreadsheet to verify the variable names for the year you are interested in.

```{r, results='hide'}
NDI_variables<-c("B23025_005", # males/females unemployed
                "C24030_019", #males in management
                "C24030_018", #males in professional occupations
                "C24030_002", #total males employed
                "B17012_002", #households in poverty
                "B17012_001", #total households - poverty determined
                "B06009_002", #less than HS education
                "B06009_001", #total with education determined, 25 years and older
                "B23025_002", #total in the labor force
                "B23025_001", #total population 16 and older
                "B25014_007", #owner-occupied housing 2+ ppl per room
                "B25014_013", #renter-occupied housing, 2+ people per room
                "B25014_001", #occupied housing units
                "B11012_010", #female-householder, own children < 18 years
                "B11012_001", #Total househods
                "B19001_006", #income 25-29.9k
                "B19001_005", #income 20-24.9k
                "B19001_004", #income 15-19.9k
                "B19001_003", #income 10-14.0k
                "B19001_002", #income 5-9.9k
                "B19001_001", #income total income
                "B19058_002", #total residents public assistance
                  "B19058_001") #total households- assistance determined
acsvars<-read_csv('ACMT/ACSColumns.csv')
acsvars<-subset(acsvars, acs_col %in% NDI_variables)
##create 'count' versions of each variable name and 'proportion' versions for each #ACS variable where applicable
acs_count_names<-paste(acsvars$var_name, "count", sep="_")
  if (length(acsvars$var_name[acsvars$universe_col != ""]) == 0) { # prevent having something that is exactly acs_proportion_names <- character(0)
  } else {
acs_proportion_names <- paste(acsvars$var_name[!is.na(acsvars$universe_col)], "proportion", sep="_") }

#set year and radius and variable names
codes_of_acs_variables_to_get<-acsvars$acs_col
names_of_variable_to_get<-c(acs_count_names, acs_proportion_names)
radius<-1000
year <- 2019

#add columns to dataset to add variables to
var.cols<-data.frame(matrix(nrow=nrow(seattle_alcohol_dispensary_list), ncol=length(names_of_variable_to_get)))
colnames(var.cols)<-names_of_variable_to_get #name the columns
seattle_alcohol_dispensary_list<-cbind(var.cols, seattle_alcohol_dispensary_list) #bind the columsn to the

```

## 4. Create a function and loop to pull ACMT measures

Once the variables are designated, we can write a loop to pull the designated variables for each location in our dataset

```{r acmt loop, results='hide'}

#run loop to pull variables
for(address in 1:nrow(seattle_alcohol_dispensary_list)) {
   tryCatch({if(!is.na(seattle_alcohol_dispensary_list[,1][address])) next #skip the row if the data is already there
  if(!is.na(seattle_alcohol_dispensary_list[,1][address])) next #skip the row if the data is already there
  print(address) #print the number to keep track of progress
  latitude<-seattle_alcohol_dispensary_list$lat[address] #set lat
  longitude<-seattle_alcohol_dispensary_list$long[address] #set long
  
  environmental_measures<-get_acmt_standard_array(long=longitude, lat=latitude, radius_meters = radius, year=year, codes_of_acs_variables_to_get = codes_of_acs_variables_to_get) #pull measures for given lat & long
 
      for(name_of_variable in names_of_variable_to_get){ #for each measures, get the value and put it into the column of the same name
     value_of_variable <- environmental_measures[environmental_measures$names == name_of_variable, ]$values  
     seattle_alcohol_dispensary_list[[name_of_variable]][address]<-value_of_variable
  }

 for (name_of_variable in names_of_variable_to_get) {
        seattle_alcohol_dispensary_list[[name_of_variable]][address] <- environmental_measures[environmental_measures$names == name_of_variable, ]$values  
 }},error=function(e){cat("ERROR :", conditionMessage(e), "\n")}) #this will print any error messages
}

```

## 5. Run a principal component analysis to find weights for each variable measure
Now we can use the ACS measures that were just pulled to calculate the NDI score for the neighborhood around each dispensary. First we need to combine measures to calculate the total percent of residents with professional degrees and the total percent of housing that is crowded (2.01 or greater occupants per room), and total percent of households with an income less than $30,000.

To create the composite NDI value, we first must get the weight for individual measure by conducting a principal component analysis of the eight designated NDI variables: (1) percent of males in professional occupations, (2) percent of households in poverty, (3) percent of residents with no high school diploma, (4) the percent of residents who are unemployed, (5) the percent of crowded housing, (6) the percent of households with income less than $30,000, (7) the percent of female-headed households with dependents less than 18 years old, and (8) the percent of households on public assistance. The loadings generated by the PCA will be used to weight each measure in calculating the overall NDI summary measure for each tract. The resulting composite NDI measure is multiplied by -1 so that a higher NDI measure indicates a neighborhood with a higher level of deprivation. Finally, we standardize the NDI measures by subtracting the mean and dividing by the standard deviation so that the measure has a mean of 0 and a standard deviation of 1.

```{r, results='hide'}
#clean-up NDI measures
seattle_dispensary_ndi<-seattle_alcohol_dispensary_list %>%
  mutate(total_males_management_professional = males_in_management_count+males_in_professional_occup_count, 
         total_crowded_housing = owner_2.01_or_more_per_room_count
         + renter_2.01_or_more_per_room_count, 
         total_income_below_30k = household_income_25_29k_count + household_income_20_24k_count + houseohld_income_15_19k_count + household_income_10_14k_count+hhincome_less_than_10k_count) %>% 
  mutate(percent_males_management_professional = total_males_management_professional / males_16_older_workforce_count, 
         percent_crowded_housing = total_crowded_housing / (total_occupied_housing_units_room_count), 
         percent_income_below_30k = total_income_below_30k/households_income_determined_count) 

#PCA Factor Analysis
NDI_measures<-seattle_dispensary_ndi %>%
  filter(!is.na(percent_males_management_professional))%>%
  subset(select=c(percent_males_management_professional, households_in_poverty_proportion, no_hsdiploma_proportion, unemployed_proportion, percent_crowded_housing, percent_income_below_30k, female_head_kids_proportion, public_asst_proportion))

ndi_pca<-princomp(NDI_measures, cor=TRUE)
```

## 6. Calculate the NDI measure
Once the NDI score has been calculated and standardized, we can construct a boxplot to compare the mean and distribution of NDI scores for establishments with a marijuana retailer license to those with an off-premise alcohol license.

```{r, results='hide'}
##assign loading values for each variable
percent_males_management_professional_loading<-ndi_pca$loadings[1]
households_in_poverty_proportion_loading<-ndi_pca$loadings[2]
no_hsdiploma_proportion_loading<-ndi_pca$loadings[3]
unemployed_proportion_loading<-ndi_pca$loadings[4]
percent_crowded_housing_loading<-ndi_pca$loadings[5]
percent_income_below_30k_loading<-ndi_pca$loadings[6]
female_head_kids_proportion_loading<-ndi_pca$loadings[7]
public_asst_proportion_loading<-ndi_pca$loadings[8]

#Calculated & standardize weighted NDI value using pca loadings
seattle_dispensary_ndi <-seattle_dispensary_ndi%>%
  mutate(ndi_value=((percent_males_management_professional*percent_males_management_professional_loading)+
           (households_in_poverty_proportion*households_in_poverty_proportion_loading)+
           (no_hsdiploma_proportion*no_hsdiploma_proportion_loading)+
           (unemployed_proportion*unemployed_proportion_loading)+
           (percent_crowded_housing*percent_crowded_housing_loading)) + 
           (percent_income_below_30k*percent_income_below_30k_loading) + 
           (female_head_kids_proportion*female_head_kids_proportion_loading)+
           (public_asst_proportion*public_asst_proportion_loading)*-1) %>%
  mutate(ndi_standardized=(ndi_value-mean(ndi_value))/sd(ndi_value)) %>%
  mutate(dispensary_active=ifelse(status=="ACTIVE (ISSUED)" | status=="PENDING (ISSUED)", 1, 0))
```

## 7. Create a boxplot to compare the NDI measures for dispensaries versus alcohol retailers
Once the NDI score has been calculated and standardized, we can construct a boxplot to compare the mean and distribution of NDI scores for establishments with a marijuana retailer license to those with an off-premise alcohol license.

```{r}

#look at specific types of alcohol privileges
boxplot(ndi_standardized~license_type,
        data=seattle_dispensary_ndi,
        main = "Neighborhood Deprivation Scores", 
        xlab = "Standardized NDI score", 
        ylab = "Type of License", 
        col = "cadetblue3", 
        border= "coral4",
        horizontal=TRUE)
```

## 8. Calculate a t-test to statistically compare NDI measures
We can also statistically compare the NDI measures of neighborhoods around alcohol and marijuana retailers using an independent t-test.

```{r}

ndi_ttest<-t.test(seattle_dispensary_ndi$ndi_value~seattle_dispensary_ndi$license_type, var.equal=TRUE)
ndi_ttest

```

## Additional boxplot
In addition to the comparisons of marijuana and alcohol retailer NDI measures, we can also compare across multiple types of alcohol retailers by first constructing a boxplot by license type for alcohol retailers.

```{r}

ndi_privilege_boxplot<- ggplot(seattle_dispensary_ndi[seattle_dispensary_ndi$license_type=='alcohol_offpremise',], aes(x=factor(privilege), y=ndi_standardized, fill=privilege))+  
  geom_boxplot()+
  coord_flip()+
  labs(title="Neighborhood Deprivation Scores", y="Standardized NDI value", x="Privilege Type", fill="Privilege Type")+
  theme_minimal()+
  theme(legend.position = 'bottom')

ndi_privilege_boxplot
```

## Additional statistical analysis
We can also look at statistical differences between the deprivation level of neighborhoods around each alcohol establishment using an ANOVA test. First we will look at the mean and sd for NDI scores by privilege type.

```{r}
#look at table of means:
ndi_mean_table<-seattle_dispensary_ndi[seattle_dispensary_ndi$license_type=='alcohol_offpremise',] %>%
  group_by(privilege) %>%
  summarise_at(vars(ndi_standardized), list(NDI_mean=mean, NDI_SD=sd))

ndi_aov<-aov(seattle_dispensary_ndi$ndi_standardized~seattle_dispensary_ndi$privilege, data=seattle_dispensary_ndi[seattle_dispensary_ndi$license_type=='alcohol_offpremise',])
print(ndi_mean_table)
```

```{r}
summary(ndi_aov)
```
## References

Messer, L.C., Laraia, B.A., Kaufman, J.K., Eyster, J., Holzman, C., Culhane, J., Elo, I., Burke, J.G., & O'Campo, P. (2006). The Development of a Standardized Neighborhood Deprivation Index. Journal of Urban Health, 84(6): 1041-1062.


