---
title: 'ACMT Example: Population density over distance'
author: "Weipeng Zhou"
date: "2/22/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit this file with the working directory of "workspace/" instead of "workspace/examples"
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = '../')
```

## Introduction
Automatic Context Measurement Tool (ACMT) is a convenient tool for studying neighbourhoods in the United States. Based on the user-provided address and radius, ACMT locates a geographical area and outputs context measurements (population, education level, commute time, etc.) for the area. ACMT is easy to install, highly reproducible and works consistently across computer platforms.

There are various ways of using ACMT and here we show one example -- 5-city comparison of population density decay as function of distance from City Hall.

We use ACMT to get population densities for 5 cities (Seattle, Los Angeles, Chicago, New York and Boston) over 5 radiuses (1000, 2000, 3000, 4000, 5000). We consider the center of a city to be the location of its City Hall. After getting the densities, we make a plot of density over radius, categoried by cities; and we will be able to tell which city is the most populated across space.

We take the following steps:

1. Find the addresses of each city's City Hall  

2. Use ACMT's geocoder to convert the addresses to coordinates  

3. Use ACMT to get the population measurement for each coordinate over 5 radiuses (1000, 2000, 3000, 4000, 5000)  

4. Calculate population density  

5. Plot density vs. radius, categoried by cities

## Example
### 1. Find the addresses of each city's City Hall
We have gathered the City Hall addresses for Seattle, Los Angeles, Chicago, New York and Boston from Google.

```{r, results = 'hide'}
source("setup-acmt.R")

library(ggplot2)
city_hall_to_address_list <- list(
  seattle_city_hall="600 4th Ave, Seattle, WA 98104",
  los_angeles_city_hall="200 N Spring St, Los Angeles, CA 90012",
  chicago_city_hall="121 N LaSalle St, Chicago, IL 60602",
  new_york_city_hall="City Hall Park, New York, NY 10007",
  boston_city_hall="1 City Hall Square #500, Boston, MA 02201"
)
```

### 2. Use ACMT's geocoder to convert the addresses to coordinates
ACMT comes with a handy geocoder that converts addresses to latitude/longtitude coordinates. We check if geocoder is available in the version of ACMT you installed; if it is not available, we use the pre-computed coordinates.

```{r, results='hide'}
convert_address_to_lat_long <- function (city_hall_to_address_list) { # function to get get lat/long for each address
  city_hall_to_lat_long_list <- vector(mode="list", length=length(city_hall_to_address_list))
  names(city_hall_to_lat_long_list) <- names(city_hall_to_address_list)
  for (name in names(city_hall_to_address_list)){
    city_hall_to_lat_long_list[[name]] <- geocode(city_hall_to_address_list[[name]])
  }
  return(city_hall_to_lat_long_list)
}
```

```{r, results='hide'}
# is geocoder available?
geocoder_is_available <- FALSE
tryCatch({  # note that codes in try is not inside a new function, just treat it as normal R code
  geocode("1 City Hall Square #500, Boston, MA 02201")
  geocoder_is_available <- TRUE
}, error = function(condition) {  # note that in error handler, the codes is inside a new function; this is why we need <<- for assigning acs_var_names
  print(condition$message)
  print("Geocoder not available: using stored address to lat/long mappings instead")
})

city_hall_to_lat_long_list <- NULL
# call geocoder if available, use hard coded info otherwise
if (geocoder_is_available) {
  city_hall_to_lat_long_list <- convert_address_to_lat_long(city_hall_to_address_list)
} else {
  city_hall_to_lat_long_list <- list(
    seattle_city_hall=list(latitude=47.60328, longitude=-122.3302),
    los_angeles_city_hall=list(latitude=34.05397, longitude=-118.2436),
    chicago_city_hall=list(latitude=41.88334, longitude=-87.63229),
    new_york_city_hall=list(latitude=40.66392, longitude=-73.93835),
    boston_city_hall=list(latitude=42.35773, longitude=-71.05919)
  )
}
```

```{r}
print(city_hall_to_lat_long_list[1:2])
```

### 3. Use ACMT to get the population measurement for each coordinate over 5 radiuses (1000, 2000, 3000, 4000, 5000)
We create a function for querying ACMT measurements for a list of coordinates and radiuses. We are interested in the variable `total_pop_count`.

```{r, results='hide'}
# function to get the environmental measures for the variables we are intersted
get_variable_measures_from_acmt <- function (city_hall_to_lat_long_list, radius_vector, year, names_of_variable_to_get, external_data_name_to_info_list=NULL, codes_of_acs_variables_to_get=NA) {
  city_hall_to_radius_to_variable_to_measures_list <- vector(mode="list", length=length(city_hall_to_lat_long_list))
  names(city_hall_to_radius_to_variable_to_measures_list) <- names(city_hall_to_lat_long_list)

  for(city_hall in names(city_hall_to_radius_to_variable_to_measures_list)) {
    radius_to_variable_to_measures_list <- vector(mode="list", length=length(radius_vector))
    names(radius_to_variable_to_measures_list) <- as.character(radius_vector)
    for (radius in radius_vector) {
      print(city_hall)
      print(radius)
      # get lat/long
      latitude <- city_hall_to_lat_long_list[[city_hall]]$latitude
      longitude <- city_hall_to_lat_long_list[[city_hall]]$longitude

      # get environmental measures for all variables
      environmental_measures <- get_acmt_standard_array(long=longitude, lat=latitude, radius_meters = radius, year=year, external_data_name_to_info_list=external_data_name_to_info_list, codes_of_acs_variables_to_get=codes_of_acs_variables_to_get)

      # get environmental measures for the variables are interested
      variable_to_measures_list <- vector(mode="list", length=length(names_of_variable_to_get))
      names(variable_to_measures_list) <- names_of_variable_to_get
      for (name_of_variable in names_of_variable_to_get) {
        value_of_variable <- environmental_measures[environmental_measures$names == name_of_variable, ]$values  # 66370.01, seattle, r=2000, year=2017
        variable_to_measures_list[[name_of_variable]] <- value_of_variable
      }
      #radius_to_variable_to_measures_list[[which(radius == radius_vector)]] <- variable_to_measures_list
      radius_to_variable_to_measures_list[[as.character(radius)]] <- variable_to_measures_list
    }
    city_hall_to_radius_to_variable_to_measures_list[[city_hall]] <- radius_to_variable_to_measures_list
  }
  return(city_hall_to_radius_to_variable_to_measures_list)
}
```

```{r, results='hide'}
city_hall_to_lat_long_list <- city_hall_to_lat_long_list
radius_vector <- c(1000, 2000, 3000, 4000, 5000)
year <- 2017
names_of_variable_to_get <- c("total_pop_count")  # the name of the variable in ACMT's returned result
codes_of_acs_variables_to_get <- c("B01001_001")  # speed up; ask ACMT to only query this variable from the ACS server to speed up computation; B01001_001 is population; check ACSColumns.csv for mappings between codes and variables.

start_time_get_variable_measures_from_acmt <- Sys.time()
city_hall_to_radius_to_variable_to_measures_list <- get_variable_measures_from_acmt(city_hall_to_lat_long_list=city_hall_to_lat_long_list, radius_vector=radius_vector, year=year, names_of_variable_to_get=names_of_variable_to_get, codes_of_acs_variables_to_get=codes_of_acs_variables_to_get)
end_time_get_variable_measures_from_acmt <- Sys.time()
```

```{r}
print(city_hall_to_radius_to_variable_to_measures_list[1])
print("Between start and end of getting ACMT measures: ")
print(end_time_get_variable_measures_from_acmt - start_time_get_variable_measures_from_acmt)
```

### 4. Calculate population density
We create a function for computing population density for the given radiues.

```{r, results = 'hide'}
add_density_measures <- function(city_hall_to_radius_to_variable_to_measures_list) {
  city_hall_to_radius_to_variable_to_measures_with_population_density_list <- city_hall_to_radius_to_variable_to_measures_list  # R's way of making a deep copy
  for (city_hall in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list)){
    for(radius_character in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]])){
      for(variable in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]][[radius_character]])) {
        if(variable == "total_pop_count"){  # add other "if" statements if you want to compute other measures
          total_pop_count <- city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]][[radius_character]][[variable]]
          radius_numeric <- as.numeric(radius_character)
          population_density <-  total_pop_count/(pi*radius_numeric^2)
          city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]][[radius_character]][["population_density"]] <- population_density
        }
      }
    }
  }
  return(city_hall_to_radius_to_variable_to_measures_with_population_density_list)
}
```

```{r, results = 'hide'}
city_hall_to_radius_to_variable_to_measures_with_population_density_list <- add_density_measures(city_hall_to_radius_to_variable_to_measures_list)
```

```{r}
print(city_hall_to_radius_to_variable_to_measures_with_population_density_list[1])
```

### 5. Plot density vs. radius, categoried by cities
We create a function for converting the data we have so far to `ggplot` friendly format. We then use `ggplot` to create our plot.

```{r, results = 'hide'}
convert_to_dataframe_for_plotting <- function (city_hall_to_radius_to_variable_to_measures_with_population_density_list) {
  city_hall_vector <- c()
  radius_vector <- c()
  variable_vector <- c()
  value_vector <- c()

  for(city_hall in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list)){
    for(radius_character in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]])){
      for(variable in names(city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]][[radius_character]])){
        city_hall_vector <- c(city_hall_vector, city_hall)
        radius_vector <- c(radius_vector, as.numeric(radius_character))
        variable_vector <- c(variable_vector, variable)
        value_vector <- c(value_vector, city_hall_to_radius_to_variable_to_measures_with_population_density_list[[city_hall]][[radius_character]][[variable]])
      }
    }
  }

  dataframe_for_plotting <- data.frame(city_hall_vector, radius_vector, variable_vector, value_vector, stringsAsFactors=FALSE)
  names(dataframe_for_plotting) <- c("city_hall", "radius", "variable", "value")
  return(dataframe_for_plotting)
}
```

```{r, results = 'hide'}
dataframe_for_plotting <- convert_to_dataframe_for_plotting(city_hall_to_radius_to_variable_to_measures_with_population_density_list)
```

```{r}
print(head(dataframe_for_plotting))
```

```{r}
ggplot(dataframe_for_plotting[dataframe_for_plotting$variable == "population_density", ], aes(x=radius)) +
        geom_line(aes(y=value, col=city_hall)) +
        labs(y="population density (people/m^2)",
             x="radius (m)")
```

## Results
We see that New York City has a much higher population density, and its density does not reduce much over radiuses.

However, this approach has a critical flaw. The population density is calculated by dividing the total population by a circular area. By doing so, we assume households are eventually distributed within the circular area. Nevertheless, this assumption can be problematic. For example, there is a significant amount of sea area near Seattle City Hall, and households are mostly in the land area. Thus, when calculating population density, the denominator should only be the land part of the circular area.

The problem is, ACMT does not naively provide land area and we need to import external data into ACMT.

In another example, we refine our analysis on population density and demonstrate how external data can be imported into ACMT. We show an example of importing National Walkability Index dataset, which includes land area data, into ACMT. We also include national walkability index in our analysis.

```{r, include=FALSE}
##Lastly, we ensure that you are reproducing the results the same as what's on other platforms.
## tests
library(testthat)
expect_equal(filter(dataframe_for_plotting[dataframe_for_plotting$variable == "population_density", ], city_hall=="seattle_city_hall", radius==2000)$value, 0.005230759)
expect_equal(filter(dataframe_for_plotting[dataframe_for_plotting$variable == "population_density", ], city_hall=="los_angeles_city_hall", radius==3000)$value, 0.005561481459)
expect_equal(filter(dataframe_for_plotting[dataframe_for_plotting$variable == "population_density", ], city_hall=="chicago_city_hall", radius==1000)$value, 0.009131229584)
expect_equal(filter(dataframe_for_plotting[dataframe_for_plotting$variable == "population_density", ], city_hall=="boston_city_hall", radius==5000)$value, 0.005571771875)
```